---
title: "Tidyverse Tutorial"
author: "Lauren Oey"
date: "4/9/2018"
output: pdf_document
---

# WELCOME TO THE TIDYVERSE

## ecosystem - collection of awesome, consistent packages
## "tidy data" philosophy
### 1. each variable forms a column
### 2. each observation forms a row
### 3. each type of observational unit forms a table
## caveat: not a replacement to R but more a really easy to use supplement

## Some Resources:
### Tidyverse: https://www.tidyverse.org
### R for Data Science by Hadley Wickham & Garrett Grolemund (online version): http://r4ds.had.co.nz

# 
```{r loadPackages}
# If you haven't installed these packages already:

# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("tidyr")
# install.packages("readr")

# tidyverse 
library(tidyverse)
detach(package:tidyverse)

library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)
```

# readr: Reading in a File
```{r}
# base R
start.time = proc.time()
oscars = read.csv("http://www.stat.columbia.edu/~gelman/arm/examples/academy.awards/oscars.csv", header = T)
(total.time.base = proc.time() - start.time)

# readr
start.time = proc.time()
oscars = read_csv("http://www.stat.columbia.edu/~gelman/arm/examples/academy.awards/oscars.csv")
(total.time.readr = proc.time() - start.time)

# Benefits of readr:
## 1. Faster
## 2. Smart! Knows that the header is meant to be the header
## 3. Doesn't convert strings to factors
```

# tidyr: Reshaping Data

## Hadley Wickham on "Tidy Data": http://vita.had.co.nz/papers/tidy-data.pdf
```{r}
# Let's start by looking at the data
# install.packages("reshape2")
library(reshape2)
head(french_fries, 10)

# Wide Format to Long/Narrow Format
(french_fries.long <- gather(french_fries, key="flavour", value="rating", 5:9))

# Long/Narrow Format to Wide Format
spread(data=french_fries.long, key="flavour", value="rating")

# Unite multiple columns into a single column
(french_fries.unite <- unite(data=french_fries, col="subj_rep", subject, rep))

# Separate single column into multiple columns
separate(french_fries.unite, subj_rep, c("subject", "rep"))
```

# dplyr: Data Wrangling
```{r}
# install.packages("wordbankr")
library(wordbankr)

# dataset on children with regard to word learning
wordz <- get_administration_data()
head(wordz,10)
summary(wordz)

# making language and form factors
wordz$language <- as.factor(wordz$language)
wordz$form <- as.factor(wordz$form)
summary(wordz)

# dataset on the word items
itemz <- get_item_data()
head(itemz, 10)


# filter() - choose certain rows (subset observations)
filter(wordz, language == "English (American)")
filter(wordz, language == "English (American)" & age <= 12)
filter(wordz, language != "English (American)" & age <= 12)
filter(wordz, language == "English (American)" | age <= 12)
filter(wordz, xor(language == "English (American)", age <= 12))

# select() - choose certain columns (subset variables)
select(wordz, c("age", "sex", "comprehension", "production"))

# arrange() - rearrange order of rows
arrange(wordz, age)
arrange(wordz, -age)
arrange(wordz, desc(age))
arrange(wordz, age, comprehension)

# mutate() - creating new variables (new columns)
mutate(wordz, prod_comp_ratio = production/comprehension)
mutate(wordz, age_sex = paste(age, sep="_", sex))
mutate(wordz, over1yr = ifelse(age >= 12, TRUE, FALSE))

# summarise() - summarizing data; collapsing data across rows into a single row
summarise(wordz, mean=mean(comprehension))
summarise(wordz, mean=mean(comprehension), sd=sd(comprehension))
```

# Practice! (Wickham, 2014)
```{r}
flights <- read_csv("flights.csv")
head(flights, 10)
# Use the flights dataframe to answer the following questions:
# 1. How many flights are there to SFO or OAK?
# 2. Order the flights by most departure delay, followed by most arrival delay.
# 3. Compute the speed. Which flight was the fastest?
```

```{r}
# 1.
nrow(filter(flights, dest %in% c("SFO","OAK")))

# 2.
arrange(flights, -dep_delay, -arr_delay)

# 3.
arrange(mutate(flights, speed = dist/time), -speed)
```

# Combining dataframes
```{r}
# left_join() / right_join() / inner_join() / full_join() - combining columns from dataframes
(a <- data.frame(col1=c("A", "B", "C"), col2=c(1,2,3)))
(b <- data.frame(col1=c("A", "B", "D"), col3=c(T,F,T)))
left_join(a,b,by="col1")
right_join(a,b,by="col1")
inner_join(a,b,by="col1")
full_join(a,b,by="col1")

semi_join(a,b,by="col1")
anti_join(a,b,by="col1")

# bind_rows() - append rows to dataframe
(c <- data.frame(col1=c("D", "E", "F"), col2=c(4,5,6)))
bind_rows(a,c)

# bind_cols() - append columns to dataframe
bind_cols(a, b)
```

# Meet the Pipe %>% : Combining Functions
```{r}
# The pipe %>% allows us to combine functions
# Object before %>% gets passed as the 1st argument in the function after %>%
# instead of declaring the dataframe, the data becomes whatever was before the pipe

wordz %>%
  filter(language == "English (American)")
# is equivalent to:
filter(wordz, language == "English (American)")

# Similarly,
wordz %>%
  filter(language == "English (American)") %>%
  arrange(age)
# is equivalent to:
arrange(filter(wordz, language == "English (American)"), age)

# The %>% makes the data more readable and more organized!

# One more function:
# group_by() - groups data into rows with same value of variable
wordz %>%
  group_by(language)
# Not much to be seen here, but...

wordz %>%
  group_by(language) %>%
  summarise(mean_comp = mean(comprehension),
            mean_prod = mean(production))

wordz %>%
  group_by(language, age) %>%
  summarise(n = n())


# ungroup() - remove previous groupings
wordz %>%
  group_by(sex, language) %>%
  mutate(mean_comp = mean(comprehension), 
         mean_prod = mean(production)) %>%
  ungroup() %>%
  group_by(sex) %>%
  summarise(newMean = mean(mean_comp))
```

# More Practice! (Wickham, 2014)
```{r}
# 4. Which destinations have the highest average delays?
flights %>%
  filter(!is.na(dep_delay)) %>%
  group_by(dest) %>%
  summarise(mean = mean(dep_delay)) %>%
  arrange(-mean)

arrange(summarise(group_by(flights, dest), mean=mean(dep_delay)), -mean)
```

# ggplot2: Data Visualization

## One discrete variable
```{r}
# Bar Plot
ggplot(wordz, aes(x=age)) +
  geom_bar() +
  ggtitle("Bar Plot: Count of Child Participants by Age")

plot(wordz$age, main="Bar Plot: Count of Child Participants by Age")
```

## One continuous variable
```{r}
ggplot(wordz, aes(x=comprehension)) +
  geom_histogram(binwidth=50) +
  ggtitle("Histogram: Distribution of Comprehension Scores")

ggplot(wordz, aes(x=comprehension)) +
  geom_histogram(aes(y=..density..), fill="red", colour="black", alpha=0.5) +
  geom_density() +
  ggtitle("Density Plot: Distribution of Comprehension Scores")
```

# Combining dplyr + ggplot2
## One discrete + one continuous variable
```{r}

wordz %>%
  filter(language == "English (American)") %>%
  ggplot(aes(x=as.factor(age), y=comprehension)) +
    geom_boxplot() +
    ggtitle("Boxplot: Comprehension Scores by Age for American English")

wordz %>%
  filter(language == "English (American)" & !is.na(sex)) %>%
  ggplot(aes(x=as.factor(age), y=comprehension)) +
    geom_boxplot() +
    facet_wrap(~sex) +
    ggtitle("Boxplot + Facet Wrap: Comprehension Score by Sex and Age for Am English")


wordz %>%
  filter(age < 20 & !is.na(sex) & language == "English (American)") %>%
  ggplot(aes(x=sex, y=comprehension, fill=sex)) +
  geom_violin() +
  facet_wrap(~as.factor(age)) +
  ggtitle("Violin Plot: Comprehension Score by Sex and Age (Under 20 months)") +
  theme(axis.text = element_text(size=3))


wordz %>%
  filter(age < 10 & !is.na(sex) & language == "English (American)") %>%
  ggplot(aes(x=sex, y=comprehension)) +
  geom_dotplot(binaxis="y", stackdir="center") +
  facet_wrap(~as.factor(age)) +
  ggtitle("Dot Plot: Comprehension Score by Sex and Age (Under 10 months)")


wordz %>%
  filter(language == "English (American)") %>%
  group_by(age) %>%
  summarise(mean=mean(comprehension)) %>%
  ggplot(aes(x=age, y=mean)) +
    geom_bar(stat="identity") +
    ggtitle("Bar Plot: Mean Comprehension Score by Age for American English")

wordz %>%
  filter(language == "English (American)") %>%
  group_by(age) %>%
  summarise(mean=mean(comprehension),
            sd = sd(comprehension),
            n = n()) %>%
  ggplot(aes(x=age, y=mean)) +
    geom_bar(stat="identity") +
    geom_errorbar(aes(ymin=mean-(1.96*sd)/sqrt(n), ymax=mean+(1.96*sd)/sqrt(n))) +
    ggtitle("Bar Plot w/ Error Bars: Mean Comprehension Score by Age for American English")
```

## Two continuous variables
```{r}
wordz %>%
  filter(1 != production/comprehension & age==18) %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point() +
    ggtitle("Scatter Plot: Production by Comprehension at 18 Months")

wordz %>%
  filter(1 != production/comprehension & age==18) %>%
  ggplot(aes(x=comprehension, y=production, colour=language)) +
    geom_point() +
    ggtitle("Scatter Plot w/ Colour: Production by Comprehension at 18 Months")

wordz %>%
  filter(1 != production/comprehension & age==18) %>%
  ggplot(aes(x=comprehension, y=production, colour=language)) +
    geom_point() +
    geom_rug() +
    ggtitle("Scatter Plot w/ Rug: Production by Comprehension at 18 Months")

# Fitting a model
wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point() +
    geom_smooth() +
    ggtitle("Fitting Curve: Production by Comprehension at 18 Months in American English")

wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point() +
    geom_smooth(method='lm') +
    ggtitle("Fitting Linear Model: Production by Comprehension at 18 Months")
```

## Very, very customizable
```{r}
wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production, shape=sex)) +
    geom_point(size=8, colour="purple") +
    geom_smooth(method='lm') +
    ggtitle("So Much Customizability...") +
    scale_x_continuous("Words Comprehended") +
    scale_y_continuous("Words Produced") +
    theme_bw()

wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point(shape=35, size=8, colour="purple") +
    geom_smooth(method='lm') +
    ggtitle("So Much Customizability...") +
    scale_x_continuous("Words Comprehended", limits=c(0,1000)) +
    scale_y_continuous("Words Produced", limits=c(0,350)) +
    theme_bw()

wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point(shape=35, size=8, colour="purple") +
    geom_smooth(method='lm') +
    ggtitle("So Much Customizability...") +
    scale_x_continuous("Words Comprehended", limits=c(0,1000)) +
    scale_y_reverse("Words Produced") +
    theme_bw() +
    theme(axis.title=element_text(size=20),
          axis.text=element_text(angle=180,size=15),
          plot.title=element_text(colour="red", face="bold.italic", size=20))

wordz %>%
  filter(1 != production/comprehension & age==18 & language=="English (American)") %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_point(shape=35, size=8, colour="purple") +
    geom_smooth(method='lm', color="darkgreen", fill="green") +
    ggtitle("So Much Customizability...") +
    scale_x_continuous("Words Comprehended", limits=c(0,1000)) +
    scale_y_continuous("Words Produced", limits=c(0,350)) +
    theme_minimal() +
    theme(axis.title=element_text(size=20),
          axis.text=element_text(angle=180,size=15),
          plot.title=element_text(colour="red", face="bold.italic", size=20),
          panel.background=element_rect(fill="skyblue"),
          plot.background=element_rect(fill="navyblue"))
```

## Continuous bivariate distribution
```{r}
wordz %>%
  filter(1 != production/comprehension) %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_bin2d() +
    ggtitle("Scatter Plot w/ Colour: Production by Comprehension at 18 Months")

wordz %>%
  filter(1 != production/comprehension & production < 200 & comprehension > 200 & comprehension < 400) %>%
  ggplot(aes(x=comprehension, y=production)) +
    geom_bin2d() +
    ggtitle("Scatter Plot w/ Colour: Production by Comprehension at 18 Months")
```

## Maps
```{r}
#install.package("maps")
crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)
crimesm <- reshape2::melt(crimes, id = 1)
require(maps)
states_map <- map_data("state")
ggplot(crimes, aes(map_id = state)) +
  geom_map(aes(fill = Murder), map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat)

ggplot(crimesm, aes(map_id = state)) +
  geom_map(aes(fill = value), map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  facet_wrap( ~ variable)

```


# If you want to play around with more datasets: https://vincentarelbundock.github.io/Rdatasets/datasets.html

